---
title: "02_RNApanel_clustering"
author: "SandraTietscher"
date: "2020-10-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script provides UMAP visualization of single cells and performs a first clustering step to separate tumor and stromal cells.


# Load libraries and data

First, we will load the needed libraries and the `SingleCellExperiment` object, that was created in the previous section.

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(SingleCellExperiment)
library(scater)
library(viridis)
library(RColorBrewer)
library(dittoSeq)
library(scales)
library(tidyverse)
library(ggpubr)
library(cowplot)
library(scran)

# Read SingleCellExperiment object
sce_RNA <- readRDS(paste0(getwd(), "/output/sce_RNApanel_afterQC.rds"))
```

# UMAP representation

For UMAP calculation, the arcsinh-transformed counts are used. As the goal here is only to identify main cell types, only the main cell type markers are included.
To speed up UMAP calculation and plotting, only a subset of cells is used.


## Calculate tSNE and UMAP on a subset of the cells

```{r compute-UMAP}
# Subset 10000 cells per sample
sce_RNA$cellID <- rownames(colData(sce_RNA))
  
colData(sce_RNA) %>%
  as.data.frame() %>%
  group_by(sample) %>%
  sample_n(10000) %>% 
  pull(cellID) -> cell_idx

sce_RNA_sub <- sce_RNA[,cell_idx]


# Run Dimensional reduction
markers_exclude_umap <- c("DNA1", "DNA2", "HH3", "CD206", "PD1", "Ki-67", "DapB", "GLUT1", "CCL4", "CXCL10", "CXCL9", "CCL5", "CCL18", "CXCL13_protein", "CXCL13_mRNA", "CCL22", "CCL17", "CSF1", "CD274", "CCL2", "CD11b")
markers_include_umap <- rownames(assay(sce_RNA_sub, "counts"))[! (rownames(assay(sce_RNA_sub, "counts")) %in% markers_exclude_umap)]

sce_RNA_sub <- runUMAP(sce_RNA_sub, exprs_values = "exprs", 
                     subset_row = rownames(assay(sce_RNA_sub, "exprs")) %in% markers_include_umap)

# Save subset SCE object
saveRDS(sce_RNA_sub, paste0(getwd(), "/output/RNApanel/sce_RNApanel_sub_UMAP.rds"))
```

## Visualize features on UMAP

```{r visualize-features-umap, message = FALSE, fig.width=6, fig.height=5}

# SlideID, SampleID, ROI, Indication, Tumour mask
dittoDimPlot(sce_RNA_sub, var = "sample", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_RNA)$colour_vectors$sample)

dittoDimPlot(sce_RNA_sub, var = "sample", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_RNA)$colour_vectors$sample, split.by = "sample") 

dittoDimPlot(sce_RNA_sub, var = "TIG", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_RNA)$colour_vectors$TIG)

dittoDimPlot(sce_RNA_sub, var = "TLS", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_RNA)$colour_vectors$TLS)

```

## Visualize expression on UMAP

Visualize the expression of each marker.

### 1. Arcsinh-transformed counts

```{r arcsinh-expression-umap, fig.width=35, fig.height=25}
p.list <- list()
for(i in rownames(sce_RNA_sub)){
  p.list[[i]] <- plotUMAP(sce_RNA_sub, colour_by = i, by_exprs_values = "exprs", 
                          point_size = 0.5, point_alpha = 1, shape_by = "panel_type")
}

plot_grid(plotlist = p.list, ncol = 7)
```

### 2. Scaled arcsinh-transformed counts

```{r scaled-expression-umap, message=FALSE, fig.width=35, fig.height=25}
p.list <- list()
for(i in rownames(sce_RNA_sub)){
  p.list[[i]] <- plotUMAP(sce_RNA_sub, colour_by = i, by_exprs_values = "scaled", 
                          point_size = 0.5, point_alpha = 1, shape_by = "panel_type")  + scale_colour_gradient2(name = i, low = "#2166ac", mid = "white", high = "#b2182b", limits = c(-3, 3))
}

plot_grid(plotlist = p.list, ncol = 7)
```


# Run graph-based clustering (Phenograph)

Use normalized counts for clustering, because this provides a better separation of noise and actual signal compared to arcsinh-transformed counts.  
Include only markers that are relevant for the general distinction between tumor and stromal cells

```{r phenograph-clustering}
markers_exclude_phenograph <- c("DNA1", "DNA2", "HH3", "CD206", "PD1", "Ki-67", "DapB", "GLUT1", "CCL4", "CXCL10", "CXCL9", "CCL5", "CCL18", "CXCL13_protein", "CXCL13_mRNA", "CD1c", "CD8a", "CCL22", "CCL17", "CSF1", "CD274", "CCL2", "CD11b", "SMA")
markers_include_phenograph <- rownames(assay(sce_RNA, "counts"))[! (rownames(assay(sce_RNA, "counts")) %in% markers_exclude_phenograph)]

cluster_counts_normalized <- subset(assay(sce_RNA, "normalized"), rownames(assay(sce_RNA, "normalized")) %in% markers_include_phenograph)

g <- buildSNNGraph(cluster_counts_normalized, k=8)
clust<- igraph::cluster_louvain(g)$membership
table(clust)
write.csv(table(clust), paste0(getwd(), "/output/RNApanel/cluster_cellnumbers.csv"), row.names = FALSE)

#add cluster information to sce object
sce_RNA$phenograph_cluster <- as.factor(clust)
```

## Plot mean expression for each cluster

```{r cluster-mean-expression-heatmap, message=FALSE, fig.width=6, fig.height=6}
clustermean_sce_RNA <- aggregateAcrossCells(sce_RNA, ids = DataFrame(cluster = sce_RNA$phenograph_cluster), average = TRUE)
colnames(clustermean_sce_RNA) <- clustermean_sce_RNA$cluster

# Center and scale raw counts
assay(clustermean_sce_RNA, "scaled") <- t(scale(t(assay(clustermean_sce_RNA, "counts"))))

# Normalize raw counts
norm_counts <- t(apply(assay(clustermean_sce_RNA, "counts"), 1, function(x)(x-min(x))/(max(x)-min(x))))
assay(clustermean_sce_RNA, "normalized", withDimnames = FALSE) <- norm_counts

## Plot
# Z-scaled
dittoHeatmap(clustermean_sce_RNA[markers_include_phenograph,], assay = "scaled", scale = "none",
            heatmap.colors = magma(100), 
            cluster_cols = TRUE, show_colnames = TRUE,
            breaks = seq(-3, 3, length.out = 101))

# Normalized (0-1)
dittoHeatmap(clustermean_sce_RNA[markers_include_phenograph,], assay = "normalized", scale = "none",
            heatmap.colors = viridis(100), 
            cluster_cols = TRUE, show_colnames = TRUE)

```

## Display clusters on UMAP

Use the subset SCE object for UMAP plotting.

```{r cluster-umap, fig.width=8, fig.height=7}
# Assing cluster numbers to the right cells
sub_cellnames <- colnames(sce_RNA_sub)
cluster_sub <- sce_RNA[,sub_cellnames]$phenograph_cluster
sce_RNA_sub$phenograph_cluster <- cluster_sub

#plot on umap
dittoDimPlot(sce_RNA_sub, var = "phenograph_cluster", reduction.use = "UMAP", size = 0.5, do.label = TRUE, opacity = 0.2)
dittoDimPlot(sce_RNA_sub, var = "phenograph_cluster", reduction.use = "UMAP", size = 0.5, split.by = "phenograph_cluster", opacity = 0.2)

#subset only one cluster and plot
# cells_14 <- colnames(sce_sub)[which(sce_sub$louvain_k10_norm_markersubset == "14")]
# sub.14 <- sce_sub[,cells_14]
# dittoDimPlot(sub.14, var = "louvain_k10_norm_markersubset", reduction.use = "UMAP", size = 0.5, do.label = TRUE)
```

# Assign tumor/stroma labels

Look at tumor/stroma proportion per sample.

```{r assign-TS-labels, fig.width=5, fig.height=3}
labels <- read.csv(paste0(getwd(), "/data/RNApanel_tumor_stroma_clusters.csv"))

label_list <- labels$class[match(sce_RNA$phenograph_cluster, labels$cluster)]
sce_RNA$cell_class <- label_list

#plot on UMAP
label_list_sub <- labels$class[match(sce_RNA_sub$phenograph_cluster, labels$cluster)]
sce_RNA_sub$cell_class <- label_list_sub
dittoDimPlot(sce_RNA_sub, var = "cell_class", reduction.use = "UMAP", size = 0.5, do.label = TRUE)

#Plot tumor/stroma ratio per sample
tab <- as.data.frame(prop.table(table(sce_RNA$sample, sce_RNA$cell_class), margin = 1))
colnames(tab) <- c("sample", "class", "proportion")

ggplot(tab, aes(sample, y = proportion, fill=class)) +
  geom_bar(stat="identity")+
  theme(axis.title.x=element_blank(), axis.title.y = element_blank())+
  theme(panel.background = element_blank(), axis.text.x = element_text(angle = 30))

```

# Subset tumor cell/stroma cell objects and save them

```{r subset-save-objects}
cellid <- colnames(sce_RNA)
tumor_x <- which(sce_RNA$cell_class == "tumor")
stroma_x <- which(sce_RNA$cell_class == "stroma")
cellid_tumor <- colnames(sce_RNA)[tumor_x]
cellid_stroma <- colnames(sce_RNA)[stroma_x]

sce_tumor <- sce_RNA[,cellid_tumor]
sce_stroma <- sce_RNA[,cellid_stroma]

# Save 
saveRDS(sce_tumor, paste0(getwd(), "/output/RNApanel/sce_RNApanel_tumor.rds"))
saveRDS(sce_stroma, paste0(getwd(), "/output/RNApanel/sce_RNApanel_stroma.rds"))

# Save updated objects
saveRDS(sce_RNA, paste0(getwd(), "/output/RNApanel/sce_RNApanel_TS_assigned.rds"))
saveRDS(sce_RNA_sub, paste0(getwd(), "/output/RNApanel/sce_RNApanel_sub_UMAP.rds"))
```


