---
title: "03_RNApanel_StromalSubclustering"
author: "SandraTietscher"
date: "2020-10-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script performs subclustering of stromal cells for the RNA panel. It further subclusters T/NK cells.
Finally, it merges cell type labels back into the original object, resulting in the annotated full object.

# Load libraries and data

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(SingleCellExperiment)
library(scater)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(cowplot)
library(ggpubr)
library(scran)
library(plyr)

# Read SingleCellExperiment object (with stromal cells only)
sce_stromal <- readRDS(paste0(getwd(), "/output/RNApanel/sce_RNApanel_stroma.rds"))
```

# Stromal cell subclustering

## Calculate UMAP on all stromal cells

```{r stromal-umap}
markers_exclude_umap <- c("DNA1", "DNA2", "HH3", "Ki-67", "DapB", "GLUT1", "panCK", "E-Cadherin", "CCL4", "CXCL10", "CXCL9", "CCL5", "CCL18", "CXCL13_protein", "CXCL13_mRNA", "CCL22", "CCL17", "CSF1", "CCL2", "CD11b")
markers_include_umap <- rownames(assay(sce_stromal, "counts"))[! (rownames(assay(sce_stromal, "counts")) %in% markers_exclude_umap)]

sce_stromal <- runUMAP(sce_stromal, exprs_values = "exprs", 
                     subset_row = rownames(assay(sce_stromal, "scaled")) %in% markers_include_umap)
```

### Visualize features on UMAP

```{r stromal-umap-features, fig.width=8, fig.height=8}

dittoDimPlot(sce_stromal, var = "sample", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_stromal)$colour_vectors$sample, opacity = 0.2) 

dittoDimPlot(sce_stromal, var = "sample", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_stromal)$colour_vectors$sample, split.by = "sample", opacity = 0.2) 

dittoDimPlot(sce_stromal, var = "TIG", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_stromal)$colour_vectors$TIG)

dittoDimPlot(sce_stromal, var = "TLS", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_stromal)$colour_vectors$TLS)
```

### Visualize marker expression on UMAP

```{r stromal-umap-markerexpression, fig.width=35, fig.height=25, message=FALSE, warning=FALSE}

# Plot expression of all markers
p.list <- list()
for(i in rownames(sce_stromal)){
  p.list[[i]] <- plotUMAP(sce_stromal, colour_by = i, by_exprs_values = "scaled", 
                          point_size = 0.5, point_alpha = 1, shape_by = "panel_type")  + scale_colour_gradient2(name = i, low = "#2166ac", mid = "white", high = "#b2182b", limits = c(-3, 3))
}
plot_grid(plotlist = p.list, ncol = 7)
```

## Run graph-based clustering for stromal subset (Phenograph)

Display clusters on UMAP.

```{r stromal-clustering-umap, fig.width=8, fig.height=8}
cluster_counts_stromal <- subset(assay(sce_stromal, "normalized"), rownames(assay(sce_stromal, "normalized")) %in% markers_include_umap)

g <- buildSNNGraph(cluster_counts_stromal, k=10)
clust <- igraph::cluster_louvain(g)$membership
table(clust)
write.csv(table(clust), paste0(getwd(), "/output/RNApanel/stromal_cluster_cellnumbers.csv"), row.names = FALSE)

#add cluster information to sce object
sce_stromal$stromal_cluster <- as.factor(clust)

#plot on umap
dittoDimPlot(sce_stromal, var = "stromal_cluster", reduction.use = "UMAP", size = 0.2, do.label = TRUE, opacity = 0.2)
dittoDimPlot(sce_stromal, var = "stromal_cluster", reduction.use = "UMAP", size = 0.2, split.by = "stromal_cluster", opacity = 0.2)
```

### Plot mean expression for each cluster

```{r stromal-clusters-expression-heatmap, fig.width=5, fig.height=8}
clustermean <- aggregateAcrossCells(sce_stromal, ids = DataFrame(cluster = sce_stromal$stromal_cluster), average = TRUE)
colnames(clustermean) <- clustermean$cluster

# Center and scale raw counts
assay(clustermean, "scaled") <- t(scale(t(assay(clustermean, "counts"))))

# Normalize raw counts
norm_counts <- t(apply(assay(clustermean, "counts"), 1, function(x)(x-min(x))/(max(x)-min(x))))
assay(clustermean, "normalized", withDimnames = FALSE) <- norm_counts

## Plot only relevant markers
exclude_plot <- c("DNA1", "DNA2", "HH3")
include_plot <- rownames(assay(sce_stromal, "counts"))[! (rownames(assay(sce_stromal, "counts")) %in% exclude_plot)]

## Plot
# Z-scaled
dittoHeatmap(clustermean[include_plot,], assay = "scaled", scale = "none",
            heatmap.colors = magma(100), 
            cluster_cols = TRUE, show_colnames = TRUE,
            breaks = seq(-3, 3, length.out = 101))

# Normalized (0-1)
dittoHeatmap(clustermean[include_plot,], assay = "normalized", scale = "none",
            heatmap.colors = viridis(100), 
            cluster_cols = TRUE, show_colnames = TRUE)
```

### Assign cell type labels

Manual cell type assignment based on canonical marker expression.

```{r stromal-labels, fig.width=8, fig.height=8}
labels <- read.csv(paste0(getwd(), "/data/RNApanel_stromal_clusters.csv"))

label_list <- labels$class[match(sce_stromal$stromal_cluster, labels$cluster)]
sce_stromal$celltype <- label_list

# Plot on UMAP
dittoDimPlot(sce_stromal, var = "celltype", reduction.use = "UMAP", size = 0.5, opacity = 0.5)
```

```{r stromal-proportions, fig.width=8, fig.height=6}
#Plot subtype proportions per sample
tab <- as.data.frame(prop.table(table(sce_stromal$sample, sce_stromal$celltype), margin = 1))
colnames(tab) <- c("sample", "celltype", "proportion")

ggplot(tab, aes(sample, y = proportion, fill=celltype)) +
  geom_bar(stat="identity")+
  theme(axis.title.x=element_blank(), axis.title.y = element_blank())+
  theme(panel.background = element_blank(), axis.text.x = element_text(angle = 30))
```

# T/NK cell subclustering

To cluster the T/NK population into the main subtypes, we will only use the main markers CD3, CD4, CD8 and CD7.  


## Subset T_NK cell object

Re-normalize counts after subsetting.  

```{r object-subsetting}
cellid <- colnames(sce_stromal)
TNK_x <- which(sce_stromal$celltype == "T_NK")
cellid_TNK <- colnames(sce_stromal)[TNK_x]
sce_TNK <- sce_stromal[,cellid_TNK]

#Re-normalize counts
counts <- sce_TNK@assays@data@listData$counts
perc99 <- apply(counts, 1, function(x) quantile(x, 0.99))
norm_counts <- t(apply(counts, 1, function(x)(x-min(x))/(quantile(x, 0.99)-min(x))))
norm_counts <- t(apply(norm_counts, 1, function(x) pmin(x, 1)))
assay(sce_TNK, "normalized", withDimnames = FALSE) <- norm_counts
```

## Calculate UMAP on T/NK cells
Visualize features and marker expression

```{r TNK-umap-sample}
markers_include_TNK <- c("CD3", "CD4", "CD8a", "CD7")
markers_extended <- c("CD3", "CD4", "CD8a", "CD7", "Ki-67", "CSF1", "CXCL13_mRNA", "CXCL13_protein")

sce_TNK <- runUMAP(sce_TNK, exprs_values = "normalized", 
                     subset_row = rownames(assay(sce_TNK, "normalized")) %in% markers_include_TNK)

dittoDimPlot(sce_TNK, var = "sample", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_TNK)$colour_vectors$sample)
dittoDimPlot(sce_TNK, var = "sample", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_TNK)$colour_vectors$sample, split.by = "sample")

```

```{r TNK-umap-features}

dittoDimPlot(sce_TNK, var = "TIG", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_TNK)$colour_vectors$TIG)
dittoDimPlot(sce_TNK, var = "TLS", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_TNK)$colour_vectors$TLS)
```

```{r TNK-umap-markerexpression, message=FALSE, warning=FALSE}
# Plot expression of included markers
p.list <- list()
for(i in markers_extended){
  p.list[[i]] <- plotUMAP(sce_TNK, colour_by = i, by_exprs_values = "scaled", 
                          point_size = 0.5, point_alpha = 1, shape_by = "panel_type")  + scale_colour_gradient2(name = i, low = "#2166ac", mid = "white", high = "#b2182b", limits = c(-3, 3))
}
plot_grid(plotlist = p.list, ncol = 3)

```

## T/NK Subclustering

```{r TNK-subclusters}
cluster_counts_TNK <- subset(assay(sce_TNK, "normalized"), rownames(assay(sce_TNK, "normalized")) %in% markers_include_TNK)

g <- buildSNNGraph(cluster_counts_TNK, k=50)
clust <- igraph::cluster_louvain(g)$membership
table(clust)

#add cluster information to sce object
sce_TNK$TNK_cluster <- as.factor(clust)

#plot on umap
dittoDimPlot(sce_TNK, var = "TNK_cluster", reduction.use = "UMAP", size = 0.2, do.label = TRUE)
dittoDimPlot(sce_TNK, var = "TNK_cluster", reduction.use = "UMAP", size = 0.2, split.by = "TNK_cluster")
```

### Mean expression for each T/NK subcluster

```{r TNK-subclusters-expression-heatmap, fig.width=5, fig.height=8}
clustermean <- aggregateAcrossCells(sce_TNK, ids = DataFrame(cluster = sce_TNK$TNK_cluster), average = TRUE)
colnames(clustermean) <- clustermean$cluster

# Center and scale raw counts
assay(clustermean, "scaled") <- t(scale(t(assay(clustermean, "counts"))))

# Normalize raw counts
norm_counts <- t(apply(assay(clustermean, "counts"), 1, function(x)(x-min(x))/(max(x)-min(x))))
assay(clustermean, "normalized", withDimnames = FALSE) <- norm_counts

## Plot
# Z-scaled
dittoHeatmap(clustermean[markers_extended,], assay = "scaled", scale = "none",
            heatmap.colors = magma(100), 
            cluster_cols = TRUE, show_colnames = TRUE,
            breaks = seq(-3, 3, length.out = 101))

# Normalized (0-1)
dittoHeatmap(clustermean[markers_extended,], assay = "normalized", scale = "none",
            heatmap.colors = viridis(100), 
            cluster_cols = TRUE, show_colnames = TRUE)
```

### Assign subtype labels

```{r TNK-subtype-umap, fig.width=4, fig.height=4}
labels <- read.csv(paste0(getwd(), "/data/RNApanel_TNK_clusters.csv"))

label_list <- labels$class[match(sce_TNK$TNK_cluster, labels$cluster)]
sce_TNK$T_subtype <- label_list

#plot on umap
dittoDimPlot(sce_TNK, var = "T_subtype", reduction.use = "UMAP", size = 0.5, do.label = TRUE)
```

```{r TNK-subtype-proportions, fig.width=6, fig.height=4}
#Plot subtype proportions per sample
tab <- as.data.frame(prop.table(table(sce_TNK$sample, sce_TNK$T_subtype), margin = 1))
colnames(tab) <- c("sample", "T_subtype", "proportion")

ggplot(tab, aes(sample, y = proportion, fill=T_subtype)) +
  geom_bar(stat="identity")+
  theme(axis.title.x=element_blank(), axis.title.y = element_blank())+
  theme(panel.background = element_blank(), axis.text.x = element_text(angle = 30))
```

# Combine all celltype labels into full SCE object

```{r combine-celltype-labels}
# read in objects
sce_RNA <- readRDS(paste0(getwd(), "/output/RNApanel/sce_RNApanel_TS_assigned.rds"))
sce_RNA_sub <- readRDS(paste0(getwd(), "/output/RNApanel/sce_RNApanel_sub_UMAP.rds"))


## Prepare dataframe with cell IDs and celltype labels ##

# Tumor cells
tumor_ids <- as.data.frame(colnames(sce_RNA)[which(sce_RNA$cell_class == "tumor")])
colnames(tumor_ids) <- "cellID"
tumor_ids$celltype <- "tumor"

# Stromal cells without T/NK
stromal_ids <- as.data.frame(colnames(sce_stromal)[which(sce_stromal$celltype != "T_NK")])
colnames(stromal_ids) <- "cellID"
stromal_ids$celltype <- sce_stromal[,stromal_ids$cellID]$celltype

# TNK cells
TNK_ids <- as.data.frame(colnames(sce_TNK))
colnames(TNK_ids) <- "cellID"
TNK_ids$celltype <- sce_TNK$T_subtype

# Combine
ids_table <- rbind(tumor_ids, stromal_ids, TNK_ids)

# Add to full object
cellids_ordered <- as.data.frame(colnames(sce_RNA))
colnames(cellids_ordered)<- "cellID"
cellids_merged <- join(cellids_ordered, ids_table, by = "cellID")
sce_RNA$celltype <- cellids_merged$celltype

```

## Plot celltypes on UMAP

Use the randomly subset SCE object (120.000 cells).

```{r plot-celltype-umap, fig.width=8, fig.height=8}
# Assign cluster numbers to the right cells
sub_cellnames <- colnames(sce_RNA_sub)
celltype_sub <- sce_RNA[,sub_cellnames]$celltype
sce_RNA_sub$celltype <- celltype_sub

#plot on umap
dittoDimPlot(sce_RNA_sub, var = "celltype", reduction.use = "UMAP", size = 0.5, do.label = TRUE)
```

## Plot celltype proportions per sample and per image

Includ TLS and pre-TLS images.

```{r celltype-proportion-sample, fig.width=6, fig.height=4}
# Plot cell type frequencies per sample
tab <- as.data.frame(prop.table(table(sce_RNA$sample, sce_RNA$celltype), margin = 1))
colnames(tab) <- c("sample", "celltype", "proportion")

ggplot(tab, aes(sample, y = proportion, fill=celltype)) +
  geom_bar(stat="identity")+
  theme(axis.title.x=element_blank(), axis.title.y = element_blank())+
  theme(panel.background = element_blank(), axis.text.x = element_text(angle = 30))
```

```{r celltype-proportion-image, fig.width=8, fig.height=3}
# Plot cell type frequencies per image
tab <- as.data.frame(prop.table(table(sce_RNA$ImageNumber, sce_RNA$celltype), margin = 1))
colnames(tab) <- c("image", "celltype", "proportion")

ggplot(tab, aes(image, y = proportion, fill=celltype)) +
  geom_bar(stat="identity")+
  theme(axis.title.x=element_blank(), axis.title.y = element_blank())+
  theme(panel.background = element_blank(), axis.text.x = element_text(angle = 30))
```

# Save object

```{r save-objects}
saveRDS(sce_RNA, paste0(getwd(), "/output/RNApanel/sce_RNApanel_celltypes.rds"))
```


