---
title: "08_proteinPanel_neighbourhood"
author: "SandraTietscher"
date: "2020-10-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```

This script performs pairwise neighbourhood analysis as implemented in the HistoCAT paper.  
It performs random shuffling of the cell labels in order to calculate p-values for the enrichment/depletion of pairwise spatial interactions.

# Load libraries and data

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(data.table)
library(dplyr)
library(magrittr)
library(dtplyr)
library(ggplot2)
library(parallel)
library(neighbouRhood)
library(gplots)
library(RColorBrewer)
library(corrplot)

# Read SingleCellExperiment object
sce_protein <- readRDS("output/ProteinPanel/sce_ProteinPanel_subtypes_all.rds")

# Read object relationship file (output from Cellprofiler)
dat_relation <- fread("data/cpout/Object_relationships_ProteinPanel.csv")

```

# Prepare data for neighbourhood analysis
```{r neighbouRhood-data-preparation}

# Extract the relevant information into a datatable
dat_cells <- data.table(ImageNumber = sce_protein$HIER_ImageNumber,
                        ObjectNumber = sce_protein$CellNumber,
                        label = sce_protein$subtype, 
                        group = "all")

n_perm = 1000

# Number of cores used for multicore:
ncores=10

#TLS image numbers
TLS_nr <- c(4,7,8,14,16,18,37,42,43,71,17,21,54,60,68)
nonTLS_nr <- c(1:77)
nonTLS_nr <- setdiff(nonTLS_nr, TLS_nr)

#TIG image numbers
ROI_info <- read.csv("data/ROI_info_HIER.csv")
ROI_info$ImageNumber <- c(1:77)
TIG2 <- ROI_info %>% filter(TIG == "TIG2" & TLS == "no")
TIG2 <- TIG2$ImageNumber
TIG3 <- ROI_info %>% filter(TIG == "TIG3" & TLS == "no")
TIG3 <- TIG3$ImageNumber

cols = rev(brewer.pal(11,'Spectral'))
cmap = colorRampPalette(cols)
```

# Run for all images separately and sum up number of "significant" images for each pairwise interaction

## Prepare the data and run

```{r pairwise-classic-all-run}
d = prepare_tables(dat_cells, dat_relation)

# Calculate the baseline statistics
dat_baseline = apply_labels(d[[1]], d[[2]]) %>%
  aggregate_classic()

# Calculate the permutation statistic
dat_perm = rbindlist(mclapply(1:n_perm, function(x){
  dat_labels = shuffle_labels(d[[1]])
  apply_labels(dat_labels, d[[2]]) %>%
    aggregate_classic()
},mc.cores = ncores
), idcol = 'run') 

# Calculate the p-values
dat_p <- calc_p_vals(dat_baseline, dat_perm, n_perm = 1000, p_tresh = 0.01) 
```

## Visualizations

Generate a heatmap of the number of significant interactions for the labels.

### Absolute number of images reaching significance for each pairwise interaction

```{r pairwise-classic-all, fig.width=8, fig.height=8}
# Prepare the data
pmat = dcast(dat_p, 'FirstLabel ~ SecondLabel', value.var = 'sigval', fun.aggregate = sum,
             fill=0, drop=F)

rname = pmat$FirstLabel

pmat = pmat %>%
  select(-c('FirstLabel')) %>%
  as.matrix()

row.names(pmat) <- rname

#Remove T_AllHigh
pmat <- pmat[rownames(pmat) != "T_AllHigh", colnames(pmat) != "T_AllHigh"]

# Plot the heatmap
hr <- hclust(dist(pmat), method="ward.D")
heatmap.2(pmat,
          Colv = as.dendrogram(hr),
          Rowv = as.dendrogram(hr),
          trace = "none",
          col=cmap(75),
          density.info ='none', margins=c(8,8)
)
```

### Relative number of images reaching significance for each pairwise interaction 

Here, the number of "significant" images is divided by the number of images where both of the cell types are actually present (i.e. respective cell type proportion > 0).

```{r pairwise-relative-all, fig.width=8, fig.height=8}
# Prepare data table with number of images containing both interaction partners
dat_p$present <- 1
nmat = dcast(dat_p, 'FirstLabel ~ SecondLabel', value.var = 'present', fun.aggregate = sum,
             fill=0, drop=F)

rname = nmat$FirstLabel
nmat = nmat %>%
  select(-c('FirstLabel')) %>%
  as.matrix()
row.names(nmat) <- rname

#Remove T_AllHigh
nmat <- nmat[rownames(nmat) != "T_AllHigh", colnames(nmat) != "T_AllHigh"]

# Divide number of significant images by total number of images whith cell types present
rmat <- pmat/nmat

hr <- hclust(dist(rmat), method="ward.D")
heatmap.2(rmat,
          Colv = as.dendrogram(hr),
          Rowv = as.dendrogram(hr),
          trace = "none",
          col=cmap(75),
          density.info ='none', margins=c(8,8)
)
```

# Run only for TLS images (including pre-TLS)

```{r pairwise-classic-TLs-run}
dat_cells_TLS <- dat_cells %>% filter(ImageNumber %in% TLS_nr)
dat_relation_TLS <- dat_relation %>% filter(`First Image Number` %in% TLS_nr)

d = prepare_tables(dat_cells_TLS, dat_relation_TLS)

# Calculate the baseline statistics
dat_baseline = apply_labels(d[[1]], d[[2]]) %>%
  aggregate_classic()

# Calculate the permutation statistic
dat_perm = rbindlist(mclapply(1:n_perm, function(x){
  dat_labels = shuffle_labels(d[[1]])
  apply_labels(dat_labels, d[[2]]) %>%
    aggregate_classic()
},mc.cores = ncores
), idcol = 'run') 

# Calculate the p-values
dat_p <- calc_p_vals(dat_baseline, dat_perm, n_perm = 1000, p_tresh = 0.01) 
```

## Visualizations

Generate a heatmap of the number of significant interactions for the labels.

### Absolute number of images reaching significance for each pairwise interaction

```{r pairwise-classic-TLS, fig.width=8, fig.height=8}
# Prepare the data
pmat = dcast(dat_p, 'FirstLabel ~ SecondLabel', value.var = 'sigval', fun.aggregate = sum,
             fill=0, drop=F)

rname = pmat$FirstLabel

pmat = pmat %>%
  select(-c('FirstLabel')) %>%
  as.matrix()

row.names(pmat) <- rname

#Remove T_AllHigh
pmat <- pmat[rownames(pmat) != "T_AllHigh", colnames(pmat) != "T_AllHigh"]

# Plot the heatmap
hr <- hclust(dist(pmat), method="ward.D")
heatmap.2(pmat,
          Colv = as.dendrogram(hr),
          Rowv = as.dendrogram(hr),
          trace = "none",
          col=cmap(75),
          density.info ='none', margins=c(8,8)
)
```

### Relative number of images reaching significance for each pairwise interaction 

Here, the number of "significant" images is divided by the number of images where both of the cell types are actually present (i.e. respective cell type proportion > 0).

```{r pairwise-relative-TLS, fig.width=8, fig.height=8}
# Prepare data table with number of images containing both interaction partners
dat_p$present <- 1
nmat = dcast(dat_p, 'FirstLabel ~ SecondLabel', value.var = 'present', fun.aggregate = sum,
             fill=0, drop=F)

rname = nmat$FirstLabel
nmat = nmat %>%
  select(-c('FirstLabel')) %>%
  as.matrix()
row.names(nmat) <- rname

#Remove T_AllHigh
nmat <- nmat[rownames(nmat) != "T_AllHigh", colnames(nmat) != "T_AllHigh"]

# Divide number of significant images by total number of images whith cell types present
rmat <- pmat/nmat

hr <- hclust(dist(rmat), method="ward.D")
heatmap.2(rmat,
          Colv = as.dendrogram(hr),
          Rowv = as.dendrogram(hr),
          trace = "none",
          col=cmap(75),
          density.info ='none', margins=c(8,8)
)
```


# Run only for non-TLS images

```{r pairwise-classic-nonTLs-run}
dat_cells_nonTLS <- dat_cells %>% filter(ImageNumber %in% nonTLS_nr)
dat_relation_nonTLS <- dat_relation %>% filter(`First Image Number` %in% nonTLS_nr)

d = prepare_tables(dat_cells_nonTLS, dat_relation_nonTLS)

# Calculate the baseline statistics
dat_baseline = apply_labels(d[[1]], d[[2]]) %>%
  aggregate_classic()

# Calculate the permutation statistic
dat_perm = rbindlist(mclapply(1:n_perm, function(x){
  dat_labels = shuffle_labels(d[[1]])
  apply_labels(dat_labels, d[[2]]) %>%
    aggregate_classic()
},mc.cores = ncores
), idcol = 'run') 

# Calculate the p-values
dat_p <- calc_p_vals(dat_baseline, dat_perm, n_perm = 1000, p_tresh = 0.01) 
```

## Visualizations

Generate a heatmap of the number of significant interactions for the labels.

### Absolute number of images reaching significance for each pairwise interaction

```{r pairwise-classic-nonTLS, fig.width=8, fig.height=8}
# Prepare the data
pmat = dcast(dat_p, 'FirstLabel ~ SecondLabel', value.var = 'sigval', fun.aggregate = sum,
             fill=0, drop=F)

rname = pmat$FirstLabel
pmat = pmat %>%
  select(-c('FirstLabel')) %>%
  as.matrix()
row.names(pmat) <- rname

#Remove T_AllHigh
pmat <- pmat[rownames(pmat) != "T_AllHigh", colnames(pmat) != "T_AllHigh"]

# Plot the heatmap
hr <- hclust(dist(pmat), method="ward.D")
heatmap.2(pmat,
          Colv = as.dendrogram(hr),
          Rowv = as.dendrogram(hr),
          trace = "none",
          col=cmap(75),
          density.info ='none', margins=c(8,8)
)
```

### Relative number of images reaching significance for each pairwise interaction 

Here, the number of "significant" images is divided by the number of images where both of the cell types are actually present (i.e. respective cell type proportion > 0).

```{r pairwise-relative-nonTLS, fig.width=8, fig.height=8}
# Prepare data table with number of images containing both interaction partners
dat_p$present <- 1
nmat = dcast(dat_p, 'FirstLabel ~ SecondLabel', value.var = 'present', fun.aggregate = sum,
             fill=0, drop=F)

rname = nmat$FirstLabel
nmat = nmat %>%
  select(-c('FirstLabel')) %>%
  as.matrix()
row.names(nmat) <- rname

#Remove T_AllHigh
nmat <- nmat[rownames(nmat) != "T_AllHigh", colnames(nmat) != "T_AllHigh"]

# Divide number of significant images by total number of images whith cell types present
rmat <- pmat/nmat

hr <- hclust(dist(rmat), method="ward.D")
heatmap.2(rmat,
          Colv = as.dendrogram(hr),
          Rowv = as.dendrogram(hr),
          trace = "none",
          col=cmap(75),
          density.info ='none', margins=c(8,8)
)
```

# Run only for TIG2-images (non-TLS)

```{r pairwise-classic-TIG2-run}
dat_cells_TIG2 <- dat_cells %>% filter(ImageNumber %in% TIG2)
dat_relation_TIG2 <- dat_relation %>% filter(`First Image Number` %in% TIG2)

d = prepare_tables(dat_cells_TIG2, dat_relation_TIG2)

# Calculate the baseline statistics
dat_baseline = apply_labels(d[[1]], d[[2]]) %>%
  aggregate_classic()

# Calculate the permutation statistic
dat_perm = rbindlist(mclapply(1:n_perm, function(x){
  dat_labels = shuffle_labels(d[[1]])
  apply_labels(dat_labels, d[[2]]) %>%
    aggregate_classic()
},mc.cores = ncores
), idcol = 'run') 

# Calculate the p-values
dat_p <- calc_p_vals(dat_baseline, dat_perm, n_perm = 1000, p_tresh = 0.01) 
```

## Visualizations

Generate a heatmap of the number of significant interactions for the labels.

### Absolute number of images reaching significance for each pairwise interaction

```{r pairwise-classic-TIG2, fig.width=8, fig.height=8}
# Prepare the data
pmat = dcast(dat_p, 'FirstLabel ~ SecondLabel', value.var = 'sigval', fun.aggregate = sum,
             fill=0, drop=F)

rname = pmat$FirstLabel
pmat = pmat %>%
  select(-c('FirstLabel')) %>%
  as.matrix()
row.names(pmat) <- rname

#Remove T_AllHigh
pmat <- pmat[rownames(pmat) != "T_AllHigh", colnames(pmat) != "T_AllHigh"]
```
### Relative number of images reaching significance for each pairwise interaction 

Here, the number of "significant" images is divided by the number of images where both of the cell types are actually present (i.e. respective cell type proportion > 0).

```{r pairwise-relative-TIG2, fig.width=8, fig.height=8}
# Prepare data table with number of images containing both interaction partners
dat_p$present <- 1
nmat = dcast(dat_p, 'FirstLabel ~ SecondLabel', value.var = 'present', fun.aggregate = sum,
             fill=0, drop=F)

rname = nmat$FirstLabel
nmat = nmat %>%
  select(-c('FirstLabel')) %>%
  as.matrix()
row.names(nmat) <- rname

#Remove T_AllHigh
nmat <- nmat[rownames(nmat) != "T_AllHigh", colnames(nmat) != "T_AllHigh"]

# Divide number of significant images by total number of images whith cell types present
rmat.TIG2 <- pmat/nmat

hr <- hclust(dist(rmat.TIG2), method="ward.D")
heatmap.2(rmat.TIG2,
          Colv = as.dendrogram(hr),
          Rowv = as.dendrogram(hr),
          trace = "none",
          col=cmap(75),
          density.info ='none', margins=c(8,8))

```

# Run only for TIG3-images (non-TLS)

```{r pairwise-classic-TIG3-run}
dat_cells_TIG3 <- dat_cells %>% filter(ImageNumber %in% TIG3)
dat_relation_TIG3 <- dat_relation %>% filter(`First Image Number` %in% TIG3)

d = prepare_tables(dat_cells_TIG3, dat_relation_TIG3)

# Calculate the baseline statistics
dat_baseline = apply_labels(d[[1]], d[[2]]) %>%
  aggregate_classic()

# Calculate the permutation statistic
dat_perm = rbindlist(mclapply(1:n_perm, function(x){
  dat_labels = shuffle_labels(d[[1]])
  apply_labels(dat_labels, d[[2]]) %>%
    aggregate_classic()
},mc.cores = ncores
), idcol = 'run') 

# Calculate the p-values
dat_p <- calc_p_vals(dat_baseline, dat_perm, n_perm = 1000, p_tresh = 0.01) 
```

## Visualizations

Generate a heatmap of the number of significant interactions for the labels.

### Absolute number of images reaching significance for each pairwise interaction

```{r pairwise-classic-TIG3, fig.width=8, fig.height=8}
# Prepare the data
pmat = dcast(dat_p, 'FirstLabel ~ SecondLabel', value.var = 'sigval', fun.aggregate = sum,
             fill=0, drop=F)

rname = pmat$FirstLabel
pmat = pmat %>%
  select(-c('FirstLabel')) %>%
  as.matrix()
row.names(pmat) <- rname

#Remove T_AllHigh
pmat <- pmat[rownames(pmat) != "T_AllHigh", colnames(pmat) != "T_AllHigh"]
```

### Relative number of images reaching significance for each pairwise interaction 

Here, the number of "significant" images is divided by the number of images where both of the cell types are actually present (i.e. respective cell type proportion > 0).

```{r pairwise-relative-TIG3, fig.width=8, fig.height=8}
# Prepare data table with number of images containing both interaction partners
dat_p$present <- 1
nmat = dcast(dat_p, 'FirstLabel ~ SecondLabel', value.var = 'present', fun.aggregate = sum,
             fill=0, drop=F)

rname = nmat$FirstLabel
nmat = nmat %>%
  select(-c('FirstLabel')) %>%
  as.matrix()
row.names(nmat) <- rname

#Remove T_AllHigh
nmat <- nmat[rownames(nmat) != "T_AllHigh", colnames(nmat) != "T_AllHigh"]

# Divide number of significant images by total number of images whith cell types present
rmat.TIG3 <- pmat/nmat

hr <- hclust(dist(rmat.TIG3), method="ward.D")
heatmap.2(rmat.TIG3,
          Colv = as.dendrogram(hr),
          Rowv = as.dendrogram(hr),
          trace = "none",
          col=cmap(75),
          density.info ='none', margins=c(8,8))

```

# Visualize difference between TIG2 and TIG3

Positive values indicate that TIG2 has higher enrichment/less avoidance, negative values indicate that TIG3 has higher enrichment/less avoidance.

```{r TIG-comparison, fig.width=8, fig.height=8}
rmat.delta <- rmat.TIG2-rmat.TIG3

hr <- hclust(dist(rmat.delta), method="ward.D")
heatmap.2(rmat.delta,
          Colv = as.dendrogram(hr),
          Rowv = as.dendrogram(hr),
          trace = "none",
          col=cmap(75),
          density.info ='none', margins=c(8,8))
```

