---
title: "09_RNApanel_cytokine_expression"
author: "SandraTietscher"
date: "2020-10-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script defines cytokine-expressing cells, making use of the negative control mRNA DapB.

# Load libraries and data

```{r read-libraries-and-data, message=FALSE, results="hide"}
source("code/helper_functions/detect_mRNA_expression.R")
source("code/helper_functions/validityChecks.R")
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
library(scater)
library(reshape2)
library(data.table)
library(ComplexHeatmap)
library(corrplot)
library(pheatmap)
library(grid)
library(gridExtra)
library(tidyr)
library(ggpubr)
library(cowplot)
library(Hmisc)

sce_RNA <- readRDS("output/RNApanel/sce_RNApanel_celltypes.rds")
```

# Detect Cytokine Expressing Cells

```{r cytokine-detection}
# get the names of the cytokine channels
cytokine_channels = c("CCL4", "CCL18", "CSF1", "CXCL10", "CXCL13_mRNA", "CCL2", "CCL22", "CXCL9", "CCL17", "CCL5")

# run function to define cytokine expressing cells 
output_list <- compute_difference(sce_RNA, 
                          cellID = "cellID", 
                          assay_name = "exprs", 
                          threshold = 0.01, 
                          mRNA_channels = cytokine_channels, 
                          negative_control = "DapB", 
                          return_calc_metrics = TRUE)

# overwrite SCE object
sce_RNA <- output_list$output_sce
```

## Plot results from cytokine detection

```{r detection-results-vsDapB,fig.height=15, fig.width=8}

# check difference between DapB and signal (histogram)
p.list <- list()

for(i in cytokine_channels){
  
  # subset whole data set for visualization purposes
  diff_chemo <- output_list[[i]]
  diff_chemo_sub <- diff_chemo[sample(nrow(diff_chemo), nrow(diff_chemo)*0.1), ]
  
  # significant cells defined by subtraction
  c = ggplot(data=diff_chemo_sub, aes(x=mean_negative_control, y=mean_chemokine)) + 
    geom_point(alpha=0.2, aes(col = 
                                ifelse(padj <= 0.01 & scaled_diff > 0 & mean_negative_control < 2.5, 'p<0.01', 'n.s.'))) + 
    scale_color_manual(values = c("black", "deepskyblue1")) +
    xlim(0,4) + ylim(0,4) +
    ylab(paste("Mean expression of", i, sep=" ")) +
    xlab("Mean DapB mRNAexpression") +
    ggtitle(paste("DapB mRNA vs.", i, sep = " ")) +
    theme(legend.title = element_blank())
  p.list[[i]] <- c
}

plot_grid(plotlist = p.list, ncol = 2)
```

# Basic Stats

## Basic numbers on the chemokine expressing cells 

```{r cytokine-basic-stats}
cytokines <- data.frame(colData(sce_RNA))
cytokines <- cytokines[, cytokine_channels]

# calculate the amount of cells that are positive for 1, 2 and multiple combinations. 
single_combinations = cytokines[rowSums(cytokines) == 1,]

# number of single-cytokine positive cells 
nrow(single_combinations)

double_combinations = cytokines[rowSums(cytokines) == 2,]
# number of double-cytokines positive cells 
nrow(double_combinations)


multiple_combinations = cytokines[rowSums(cytokines) >= 3,]
# number of cells that express 3 or more cytokines
nrow(multiple_combinations)


# number of double positives per cytokine
double_counts <- colSums(double_combinations)

# frequency matrix and corrplot for frequency matrix
double_combinations[double_combinations == 0] <- NA
count_matrix = psych::pairwiseCount(x=double_combinations)

# normalize the frequency matrix by the amount of double combinations that occur for each cytokine
frequency_matrix <- count_matrix
for (i in colnames(count_matrix)){
  frequency_matrix[,i] <- frequency_matrix[,i]/double_counts[i]
}
```

## Frequency double-positive cells
The next plot shows the frequencies of all double positive cell occurences. e.g. of all CXCL10 expressing cells that also express another chemokine (More than 50% express CXCL9).

```{r frequency_of_double_positive_chemokines, fig.height=10, fig.width=10}
corrplot(frequency_matrix, is.corr = FALSE, tl.col = 'black', method = 'pie', type = 'full', 
         tl.srt = 45, tl.cex = 0.8, tl.offset = 0.5, cl.length = 2, cl.cex = 1, cl.align.text = "l", cl.ratio = 0.3,
         diag=TRUE, order = "hclust")
```

## Corrplot of Frequency matrix
Now we normalize the numbers of double positives by the numbers of all respective positive chemokines. this shows that usually between 20-40 percent of chemokine expressing cells are double positive expressors.

```{r freq_corrplot}
single_counts <- colSums(single_combinations)
frequency_matrix <- count_matrix
for (i in colnames(count_matrix)){
  frequency_matrix[,i] <- frequency_matrix[,i]/single_counts[i]
}

corrplot(frequency_matrix, is.corr = FALSE, tl.col = 'black', method = 'pie', type = 'full', 
         tl.srt = 45, tl.cex = 0.8, tl.offset = 0.5, cl.length = 2, cl.cex = 1, cl.align.text = "l", cl.ratio = 0.3,
         cl.lim = c(0,1), diag=TRUE, order = "hclust")
```
## Save SCE object

```{r save sce object}
# general cytokine producer tag for every cell (logical binary)
sce_RNA$cytokine <- ifelse(rowSums(data.frame(colData(sce_RNA)[,cytokine_channels])) > 0, TRUE, FALSE)

# rename colData entry names
idx <- match(cytokine_channels, colnames(colData(sce_RNA)))
for(i in idx){
  colnames(colData(sce_RNA))[i] <- paste0(colnames(colData(sce_RNA))[i], "_expression")
}

saveRDS(object = sce_RNA,file = "output/RNApanel/sce_RNApanel_celltypes_cytokines.rds")
```

# Advanced stats

## Plot chemokine expression by celltype

```{r chemokine-celltype-proportion, fig.width=8, fig.height=7}
# Plot percentage of each cell type that expresses a certain cytokine

cytokine_prop <- data.frame("celltype" = levels(as.factor(sce_RNA$celltype)))
cytokine_colnames <- paste0(cytokine_channels, "_expression")
for (i in cytokine_colnames) {
  cytokine_prop[[i]] <- prop.table(table(sce_RNA$celltype, colData(sce_RNA)[,i]), margin = 1)[,2]
}

cytokine_prop_g <- gather(cytokine_prop, key = "cytokine", value = "proportion_expressing", 2:11)

ggplot(cytokine_prop_g, aes(cytokine, proportion_expressing, fill = cytokine))+
  geom_bar(stat="identity")+
  facet_wrap(~celltype)+
  theme(axis.ticks.x = element_blank(),
        axis.text.x=element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank(),
        legend.background = element_blank())+
  ggtitle("Proportion of celltype that expresses a given cytokine")

ggplot(cytokine_prop_g, aes(celltype, proportion_expressing, fill = celltype))+
  geom_bar(stat="identity")+
  facet_wrap(~cytokine)+
  theme(axis.ticks.x = element_blank(),
        axis.text.x=element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank(),
        legend.background = element_blank())+
   ggtitle("Proportion of celltype that expresses a given cytokine")
```

## Plot chemokine expression by celltype - for non-TLS images only

```{r chemokine-celltype-proportion-nonTLS, fig.width=8, fig.height=7}
sce_TLSex <- sce_RNA[,which(sce_RNA$TLS == "no")]
sce_TLSex$ImageNumber <- droplevels(sce_TLSex$ImageNumber)

# Plot percentage of each cell type that expresses a certain cytokine
cytokine_prop <- data.frame("celltype" = levels(as.factor(sce_TLSex$celltype)))
cytokine_colnames <- paste0(cytokine_channels, "_expression")
for (i in cytokine_colnames) {
  cytokine_prop[[i]] <- prop.table(table(sce_TLSex$celltype, colData(sce_TLSex)[,i]), margin = 1)[,2]
}

cytokine_prop_g <- gather(cytokine_prop, key = "cytokine", value = "proportion_expressing", 2:11)

ggplot(cytokine_prop_g, aes(cytokine, proportion_expressing, fill = cytokine))+
  geom_bar(stat="identity")+
  facet_wrap(~celltype)+
  theme(axis.ticks.x = element_blank(),
        axis.text.x=element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank(),
        legend.background = element_blank())

ggplot(cytokine_prop_g, aes(celltype, proportion_expressing, fill = celltype))+
  geom_bar(stat="identity")+
  facet_wrap(~cytokine)+
  theme(axis.ticks.x = element_blank(),
        axis.text.x=element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank(),
        legend.background = element_blank())
```

# Compare cytokine expression in TIGs

In non-TLS images only.

```{r TIG-comparison-all, fig.width=4, fig.height=3}
# Compare total proportion of expressing cells
tab <- as.data.frame(prop.table(table(sce_TLSex$cytokine, sce_TLSex$TIG), margin = 2))
colnames(tab) <- c("Cytokine_expressing", "TIG", "proportion")

ggplot(tab, aes(TIG, proportion))+
  geom_col(aes(fill=Cytokine_expressing))+
  scale_fill_manual(values = c("#E69F00", "#56B4E9"))+
  theme(axis.title.x = element_blank())
```

```{r TIG-comparison-byCelltype, fig.width=6, fig.height=5}

# For each celltype individually
celltypes <- levels(as.factor(sce_TLSex$celltype))
tab <- data.frame( TIG2 = double(), TIG3 = double())

for (i in celltypes) {
  sce_i <- sce_TLSex[,which(sce_TLSex$celltype == i)]
  tab[i,] <- prop.table(table(sce_i$cytokine, sce_i$TIG), margin = 2)[2,]
}

tab$celltype <- row.names(tab)
tab <- pivot_longer(tab, 1:2, names_to = "TIG", values_to = "proportion")
ggplot(tab, aes(celltype, proportion, fill = TIG))+
  geom_bar(stat = "identity", positio = "dodge")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.25),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())+
  ylab("Proportion expressing at least one cytokine")
```

## Correlation of myeloid cells that express T-attracting chemokines with T cells 
(per ROI, excluding TLS)

```{r macrophage-cytokines-CD8, fig.widt=5, fig.height=5}

# T-attracting chemokine producer tag for every myeloid cell (logical binary)
sce_myeloid <- sce_TLSex[,which(sce_TLSex$celltype == "myeloid")]
sce_myeloid$Tattr_chemokine <- ifelse(rowSums(data.frame(colData(sce_myeloid)[,c("CCL2_expression", "CCL17_expression", "CCL4_expression", "CCL5_expression", "CXCL10_expression", "CXCL9_expression")])) > 0, TRUE, FALSE)

Tattr_chemokine_img <- prop.table(table(sce_myeloid$ImageNumber, sce_myeloid$Tattr_chemokine), margin = 1)[,2]
celltype_img <- as.data.frame(table(sce_TLSex$ImageNumber, sce_TLSex$celltype))
celltype_img <- pivot_wider(celltype_img, names_from = "Var2", values_from = "Freq")
Tattr_chemokine_comb <- cbind(Tattr_chemokine_img, celltype_img[,-1])

img_TIG <- as.data.frame(table(sce_TLSex$ImageNumber, sce_TLSex$TIG)) %>% filter(Freq > 0)
rownames(img_TIG) <- img_TIG$Var1
Tattr_chemokine_comb <- cbind(Tattr_chemokine_comb, img_TIG[,"Var2"][match(rownames(Tattr_chemokine_comb), rownames(img_TIG))])
colnames(Tattr_chemokine_comb)[13] <- "TIG"

corrs <- rcorr(as.matrix(Tattr_chemokine_comb[,1:12]), type = "spearman")
r_cor <- corrs$r[1, 2:12]
p_cor <- corrs$P[1, 2:12]

# Scatterplot
ggplot(Tattr_chemokine_comb, aes(Tattr_chemokine_img, T_CD8))+
  geom_point(aes(color=TIG))+
  geom_smooth(method = lm, color = "black", se=FALSE)+
  stat_cor(method="spearman", label.x.npc = "right")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())+
  xlab("Proportion of myeloid cells expressing\nat least one T-cell attracting chemokine")+
  ylab("Number of CD8+ T cells")

ggplot(Tattr_chemokine_comb, aes(Tattr_chemokine_img, T_CD4))+
  geom_point(aes(color=TIG))+
  geom_smooth(method = lm, color = "black", se=FALSE)+
  stat_cor(method="spearman", label.x.npc = "right")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())+
  xlab("Proportion of myeloid cells expressing\nat least one T-cell attracting chemokine")+
  ylab("Number of CD4+ T cells")

ggplot(Tattr_chemokine_comb, aes(Tattr_chemokine_img, T_CD4+T_CD8))+
  geom_point(aes(color=TIG))+
  geom_smooth(method = lm, color = "black", se=FALSE)+
  stat_cor(method="spearman", label.x.npc = "right")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())+
  xlab("Proportion of myeloid cells expressing\nat least one T-cell attracting chemokine")+
  ylab("Number of T cells")
```