---
title: "10_RNApanel_TNK_cytokines"
author: "SandraTietscher"
date: "2020-10-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

This script defines CXCL13-expressing T cells and CSF1-expressing T/NK cells and looks into them in more detail.


# Load libraries and data

```{r load-libraries-and-data, message=FALSE}
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
library(scater)
library(tidyr)
library(ggpubr)
library(dittoSeq)
library(Seurat)

sce_RNA <- readRDS(file = "output/RNApanel/sce_RNApanel_celltypes_cytokines.rds")
TIG2 <- c("TBB111", "TBB129", "TBB165", "TBB171", "TBB184", "TBB338")
```

# Assign new subtypes to CXCL13-expressing CD4 and CD8 T cells

```{r}
sce_RNA$subtype <- sce_RNA$celltype
sce_RNA$subtype <- ifelse(sce_RNA$celltype == "T_CD4" & sce_RNA$CXCL13_mRNA_expression == 1, "T_CD4_CXCL13", sce_RNA$subtype)
sce_RNA$subtype <- ifelse(sce_RNA$celltype == "T_CD8" & sce_RNA$CXCL13_mRNA_expression == 1, "T_CD8_CXCL13", sce_RNA$subtype)
```

## CXCL13+ T cell proportions

### Across TLS states

```{r CXCL13-frequency-TLS}
sce_CD8 <- sce_RNA[,which(sce_RNA$celltype == "T_CD8")]
sce_CD4 <- sce_RNA[,which(sce_RNA$celltype == "T_CD4")]


# Boxplots (individual ROIs as points)
tab_CD8 <- as.data.frame(prop.table(table(sce_CD8$ImageNumber, sce_CD8$subtype), margin = 1)) %>% filter(Var2 == "T_CD8_CXCL13")
frequency <- as.data.frame(table(sce_CD8$ImageNumber, sce_CD8$CXCL13_mRNA_expression)) %>% filter(Var2 == 1)
tab_CD8$frequency <- frequency[,3]

tab_CD4 <- as.data.frame(prop.table(table(sce_CD4$ImageNumber, sce_CD4$subtype), margin = 1)) %>% filter(Var2 == "T_CD4_CXCL13")
frequency <- as.data.frame(table(sce_CD4$ImageNumber, sce_CD4$CXCL13_mRNA_expression)) %>% filter(Var2 == 1)
tab_CD4$frequency <- frequency[,3]

tab_comb <- rbind(tab_CD8, tab_CD4)
colnames(tab_comb) <- c("ImageNumber", "subtype", "proportion", "frequency")


tab <- as.data.frame(table(sce_CD8$ImageNumber, sce_CD8$TLS))
colnames(tab) <- c("ImageNumber", "TLS", "cellnumber_image")
tab <- tab %>% filter(cellnumber_image > 0)
tab$cellnumber_image <- NULL
tab_comb2 <- merge(tab_comb, tab, by = "ImageNumber")

my_comparisons <- list( c("no", "pre"), c("no", "yes"), c("pre", "yes") )

ggplot(tab_comb2, aes(TLS, frequency))+
  geom_boxplot()+
  geom_point()+
  stat_compare_means(comparisons = my_comparisons, label = "p.format", method = "wilcox")+
  facet_wrap(~subtype, scales = "fixed")+
    theme(axis.ticks.x = element_blank(),
        axis.text.x=element_text(angle = 90, hjust=1),
        axis.title.x = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank(),
        legend.background = element_blank())+
  ylab("Frequency per ROI")
```

```{r CXCL13-proportion-TLS}
ggplot(tab_comb2, aes(TLS, proportion))+
  geom_boxplot()+
  geom_point()+
  stat_compare_means(comparisons = my_comparisons, label = "p.format", method = "wilcox")+
  facet_wrap(~subtype, scales = "fixed")+
    theme(axis.ticks.x = element_blank(),
        axis.text.x=element_text(angle = 90, hjust=1),
        axis.title.x = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank(),
        legend.background = element_blank())+
  ylab("Proportion of all CD8+/CD4+ T cells per ROI")
```

### Across samples/TIGs (for non-TLS only)

```{r CXCL13-proportion-TIG}
sce_CD8_TLSex <- sce_CD8[,which(sce_CD8$TLS == "no")]
sce_CD4_TLSex <- sce_CD4[,which(sce_CD4$TLS == "no")]


# Boxplots (individual ROIs as points)
tab_CD8 <- as.data.frame(prop.table(table(sce_CD8_TLSex$sample, sce_CD8_TLSex$subtype), margin = 1)) %>% filter(Var2 == "T_CD8_CXCL13")
frequency <- as.data.frame(table(sce_CD8_TLSex$sample, sce_CD8_TLSex$subtype)) %>% filter(Var2 == "T_CD8_CXCL13")
tab_CD8$frequency <- frequency[,3]

tab_CD4 <- as.data.frame(prop.table(table(sce_CD4_TLSex$sample, sce_CD4_TLSex$subtype), margin = 1)) %>% filter(Var2 == "T_CD4_CXCL13")
frequency <- as.data.frame(table(sce_CD4_TLSex$sample, sce_CD4_TLSex$subtype)) %>% filter(Var2 == "T_CD4_CXCL13")
tab_CD4$frequency <- frequency[,3]

tab_comb <- rbind(tab_CD8, tab_CD4)
colnames(tab_comb) <- c("sample", "subtype", "proportion", "frequency")
tab_comb$TIG <- ifelse(tab_comb$sample %in% TIG2, "TIG2", "TIG3")


ggplot(tab_comb, aes(TIG, proportion))+
  geom_boxplot()+
  geom_point()+
  stat_compare_means(label = "p.format", method = "wilcox", label.x.npc = "middle")+
  facet_wrap(~subtype, scales = "fixed")+
    theme(axis.ticks.x = element_blank(),
        axis.text.x=element_text(angle = 90, hjust=1),
        axis.title.x = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank(),
        legend.background = element_blank())+
  ylab("Proportion out of all CD8+/CD4+ T cells")
```

## Check Ki-67 status

Ki-67 expression cutoff was chosen at 0.3 mean counts per cell.

```{r Ki67-status-CXCL13}
# For CD8+ CXCL13+ T cells
Ki67_exprs <- as.data.frame(assay(sce_CD8, "counts")["Ki-67",])
colnames(Ki67_exprs) <- "Ki67_exprs"
Ki67_exprs$Ki67_status <- ifelse(Ki67_exprs$Ki67_exprs > 0.3, "positive", "negative")
tab <- cbind(Ki67_exprs, colData(sce_CD8))
tab$CXCL13_status <- ifelse(tab$CXCL13_mRNA_expression == 1, "positive", "negative")

# For CD4+ CXCL13+ T cells
Ki67_exprs_CD4 <- as.data.frame(assay(sce_CD4, "counts")["Ki-67",])
colnames(Ki67_exprs_CD4) <- "Ki67_exprs"
Ki67_exprs_CD4$Ki67_status <- ifelse(Ki67_exprs_CD4$Ki67_exprs > 0.3, "positive", "negative")
tab_CD4 <- cbind(Ki67_exprs_CD4, colData(sce_CD4))
tab_CD4$CXCL13_status <- ifelse(tab_CD4$CXCL13_mRNA_expression == 1, "positive", "negative")

tab_comb <- rbind(tab, tab_CD4)

ggplot(tab_comb, aes(x = CXCL13_status, fill = Ki67_status))+
  geom_bar(position="fill")+
  scale_fill_manual(values = c("black", "deepskyblue1"))+
  facet_wrap(~celltype)+
  theme(axis.line.x = element_line(colour = "black", size = 0.25),
        axis.line.y = element_line(colour = "black", size = 0.25),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  ylab("Proportion [%]")
```

# Assign CSF1+ NKT cell subtype

All CSF1-positive T and NK cells (except CXCL13+) will be classified as NKT cells. This is only an approximation, but as close as we can get to defining this cell type.

```{r CSF1-assign-subtype}
sce_RNA$subtype <- ifelse(sce_RNA$subtype %in% c("NK", "T_CD8", "T_CD4") & sce_RNA$CSF1_expression == 1, "CSF1_NKT", sce_RNA$subtype)
```

## Check CSF1_NKT proportions/frequency (out of all cells)

### Across TLS states

```{r CSF1-frequency-TLS, fig.width=3, fig.height=4}
tab <- as.data.frame(prop.table(table(sce_RNA$ImageNumber, sce_RNA$subtype), margin = 1)) %>% filter(Var2 == "CSF1_NKT")
frequency <- as.data.frame(table(sce_RNA$ImageNumber, sce_RNA$subtype)) %>% filter(Var2 == "CSF1_NKT")
tab$frequency <- frequency[,3]
colnames(tab) <- c("ImageNumber", "subtype", "proportion", "frequency")


tab2 <- as.data.frame(table(sce_RNA$ImageNumber, sce_RNA$TLS))
colnames(tab2) <- c("ImageNumber", "TLS", "cellnumber_image")
tab2 <- tab2 %>% filter(cellnumber_image > 0)
tab2$cellnumber_image <- NULL
tab_comb <- merge(tab2, tab, by = "ImageNumber")

my_comparisons <- list( c("no", "pre"), c("no", "yes"), c("pre", "yes") )

ggplot(tab_comb, aes(TLS, frequency))+
  geom_boxplot()+
  geom_point()+
  stat_compare_means(comparisons = my_comparisons, label = "p.format", method = "wilcox")+
  facet_wrap(~subtype, scales = "fixed")+
    theme(axis.ticks.x = element_blank(),
        axis.text.x=element_text(angle = 90, hjust=1),
        axis.title.x = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank(),
        legend.background = element_blank())+
  ylab("Frequency per ROI")
```

```{r CSF1-proportion-TLS, fig.width=3, fig.height=4}
ggplot(tab_comb, aes(TLS, proportion))+
  geom_boxplot()+
  geom_point()+
  stat_compare_means(comparisons = my_comparisons, label = "p.format", method = "wilcox")+
  facet_wrap(~subtype, scales = "fixed")+
    theme(axis.ticks.x = element_blank(),
        axis.text.x=element_text(angle = 90, hjust=1),
        axis.title.x = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank(),
        legend.background = element_blank())+
  ylab("Proportion of all cells per ROI")
```
### Across samples/TIGs

For non-TLS images only.

```{r cSF1-proportion-TIG, fig.width=3, fig.height=4}
sce_TLSex <- sce_RNA[,which(sce_RNA$TLS == "no")]

# Boxplots (individual ROIs as points)
tab <- as.data.frame(prop.table(table(sce_TLSex$sample, sce_TLSex$subtype), margin = 1)) %>% filter(Var2 == "CSF1_NKT")
frequency <- as.data.frame(table(sce_TLSex$sample, sce_TLSex$subtype)) %>% filter(Var2 == "CSF1_NKT")
tab$frequency <- frequency[,3]

colnames(tab) <- c("sample", "subtype", "proportion", "frequency")
tab$TIG <- ifelse(tab$sample %in% TIG2, "TIG2", "TIG3")

ggplot(tab, aes(TIG, proportion))+
  geom_boxplot()+
  geom_point()+
  stat_compare_means(label = "p.format", method = "wilcox", label.x.npc = "middle")+
  facet_wrap(~subtype, scales = "fixed")+
    theme(axis.ticks.x = element_blank(),
        axis.text.x=element_text(angle = 90, hjust=1),
        axis.title.x = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank(),
        legend.background = element_blank())+
  ylab("Proportion out of all cells")
```

# Compare to 10x proportions
Exclude TLS and pre-TLS

## Compare CSF1_NKT proportions (out of all T/NK cells)

```{r CSF1-proportions-method-comparison}
# Read in Seurat object and calculate NKT proportions per sample
Tcells_10x <- readRDS("data/RNAseq_Tcells.rds")
prop <- as.data.frame(prop.table(table(Tcells_10x$orig.ident, Tcells_10x$metacluster), margin = 1))
colnames(prop) <- c("sample", "metacluster", "RNAseq_proportion")
prop <- prop[prop$metacluster == "NKT",]

# Remove samples that were not measured by IMC
not.measured <- c("TBB011", "TBB330")
prop <- prop[!(prop$sample %in% not.measured),]

# Retrieve IMC proportions (of all T/NK cells)
sce_TNK <- sce_RNA[, which(sce_RNA$celltype %in% c("NK", "T_CD4", "T_CD8"))]
IMC_prop <- as.data.frame(prop.table(table(sce_TNK$sample, sce_TNK$subtype), margin = 1)) %>% filter(Var2 == "CSF1_NKT")
colnames(IMC_prop) <- c("sample", "subtype", "IMC_proportion")

# Merge
prop.comb <- merge(IMC_prop, prop, by = "sample")
prop.comb$TIG <- ifelse(prop.comb$sample %in% TIG2, "TIG2", "TIG3")

# Test correlation
ggplot(prop.comb, aes(IMC_proportion, RNAseq_proportion)) +
  geom_point(aes(color=TIG))+
  geom_smooth(method = lm, color = "black", se=FALSE)+
  stat_cor(method="spearman")+
  theme(axis.line.x = element_line(colour = "black", size = 0.25),
        axis.line.y = element_line(colour = "black", size = 0.25),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())+
  ggtitle("CSF1_NKT proportion out of all T/NK cells - method comparison")

```

## Compare CD4_CXCL13 proportions (out of all T/NK cells)

```{r, CD4_CXCL13-proportions-method-comparison}
# Read in Seurat object and calculate NKT proportions per sample
prop <- as.data.frame(prop.table(table(Tcells_10x$orig.ident, Tcells_10x$metacluster), margin = 1))
colnames(prop) <- c("sample", "metacluster", "RNAseq_proportion")
prop <- prop[prop$metacluster == "CD4_exhausted",]

# Remove samples that were not measured by IMC
not.measured <- c("TBB011", "TBB330")
prop <- prop[!(prop$sample %in% not.measured),]

# Retrieve IMC proportions (of all T/NK cells)
sce_TNK <- sce_RNA[, which(sce_RNA$celltype %in% c("NK", "T_CD4", "T_CD8"))]
IMC_prop <- as.data.frame(prop.table(table(sce_TNK$sample, sce_TNK$subtype), margin = 1)) %>% filter(Var2 == "T_CD4_CXCL13")
colnames(IMC_prop) <- c("sample", "subtype", "IMC_proportion")

# Merge
prop.comb <- merge(IMC_prop, prop, by = "sample")
prop.comb$TIG <- ifelse(prop.comb$sample %in% TIG2, "TIG2", "TIG3")

# Test correlation
ggplot(prop.comb, aes(IMC_proportion, RNAseq_proportion)) +
  geom_point(aes(color=TIG))+
  geom_smooth(method = lm, color = "black", se=FALSE)+
  stat_cor(method="spearman")+
  theme(axis.line.x = element_line(colour = "black", size = 0.25),
        axis.line.y = element_line(colour = "black", size = 0.25),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())+
  ggtitle("CD4_CXCL13+ proportion out of all T/NK cells - method comparison")

```

## Compare CD8_CXCL13 proportions (out of all T/NK cells)

```{r CD8_CXCL13-proportions-method-comparison}
# Read in Seurat object and calculate NKT proportions per sample
prop <- as.data.frame(prop.table(table(Tcells_10x$orig.ident, Tcells_10x$metacluster), margin = 1))
colnames(prop) <- c("sample", "metacluster", "RNAseq_proportion")
prop <- prop[prop$metacluster == "CD8_exhausted",]

# Remove samples that were not measured by IMC
not.measured <- c("TBB011", "TBB330")
prop <- prop[!(prop$sample %in% not.measured),]

# Retrieve IMC proportions (of all T/NK cells)
sce_TNK <- sce_RNA[, which(sce_RNA$celltype %in% c("NK", "T_CD4", "T_CD8"))]
IMC_prop <- as.data.frame(prop.table(table(sce_TNK$sample, sce_TNK$subtype), margin = 1)) %>% filter(Var2 == "T_CD8_CXCL13")
colnames(IMC_prop) <- c("sample", "subtype", "IMC_proportion")

# Merge
prop.comb <- merge(IMC_prop, prop, by = "sample")
prop.comb$TIG <- ifelse(prop.comb$sample %in% TIG2, "TIG2", "TIG3")

# Test correlation
ggplot(prop.comb, aes(IMC_proportion, RNAseq_proportion)) +
  geom_point(aes(color=TIG))+
  geom_smooth(method = lm, color = "black", se=FALSE)+
  stat_cor(method="spearman")+
  theme(axis.line.x = element_line(colour = "black", size = 0.25),
        axis.line.y = element_line(colour = "black", size = 0.25),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())+
  ggtitle("CD8_CXCL13+ proportion out of all T/NK cells - method comparison")

```

# Check correlation of CSF1_NKT cells with myeloid activation/differentiation status (across ROIs)

## Use the cytokine-expression status as a proxy for cell activation

```{r myeloid-activation-cytokines-CSF1}
sce_myeloid <- sce_RNA[,which(sce_RNA$celltype == "myeloid")]

cytokine_myeloid <- prop.table(table(sce_myeloid$ImageNumber, sce_myeloid$cytokine), margin = 1)[,2]
CSF1_freq <- as.data.frame(table(sce_RNA$ImageNumber, sce_RNA$subtype)) %>% filter(Var2 == "CSF1_NKT")
colnames(CSF1_freq) <- c("ImageNumber", "subtype", "CSF1_NKT frequency")
CSF1_freq$myeloid_activation <- cytokine_myeloid

img_TIG <- as.data.frame(table(sce_myeloid$ImageNumber, sce_myeloid$TIG)) %>% filter(Freq > 0)
colnames(img_TIG) <- c("ImageNumber", "TIG", "cellnumber")
CSF1_freq <- merge(CSF1_freq, img_TIG, by = "ImageNumber")

ggplot(CSF1_freq, aes(`CSF1_NKT frequency`, myeloid_activation)) +
  geom_point(aes(color = TIG))+
  geom_smooth(method = lm, color = "black", se=FALSE)+
  stat_cor(method="spearman")+
  theme(axis.line.x = element_line(colour = "black", size = 0.25),
        axis.line.y = element_line(colour = "black", size = 0.25),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())+
  ylab("Proportion of myeloid cells that express at least one cytokine")
```

## Use the expression of CD68, CD163, CD204 and CCL18 as a proxy for cell activation

```{r myeloid-activation-score-CSF1}
# Re-normalize counts
norm_counts <- t(apply(assay(sce_myeloid, "counts"), 1, function(x)(x-min(x))/(quantile(x, 0.99)-min(x))))
norm_counts <- t(apply(norm_counts, 1, function(x) pmin(x, 1)))
assay(sce_myeloid, "normalized", withDimnames = FALSE) <- norm_counts

# For each cells, average across the normalized counts for CD68, CD163, CD204 and CCL18
count_matrix <- assay(sce_myeloid, "normalized")[c("CD68", "CD163", "CD204", "CCL18"),]
sce_myeloid$activation_score <- colMeans(count_matrix)

activation_score_mean <- aggregateAcrossCells(sce_myeloid, sce_myeloid$ImageNumber, coldata_merge = list(activation_score = mean))$activation_score
CSF1_freq <- as.data.frame(table(sce_RNA$ImageNumber, sce_RNA$subtype)) %>% filter(Var2 == "CSF1_NKT")
colnames(CSF1_freq) <- c("ImageNumber", "subtype", "CSF1_NKT frequency")
CSF1_freq$myeloid_activation <- activation_score_mean
CSF1_freq <- merge(CSF1_freq, img_TIG, by = "ImageNumber")

ggplot(CSF1_freq, aes(`CSF1_NKT frequency`, myeloid_activation)) +
  geom_point(aes(color=TIG))+
  geom_smooth(method = lm, color = "black", se=FALSE)+
  stat_cor(method="spearman")+
  theme(axis.line.x = element_line(colour = "black", size = 0.25),
        axis.line.y = element_line(colour = "black", size = 0.25),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())+
  ylab("Mean activation score for myeloid cells")
```

# Save updated SCE object

```{r}
saveRDS(sce_RNA, "output/RNApanel/sce_RNApanel_cytokines_subtypes.rds")
```




