---
title: "13_proteinPanel_neighbourhood_extended"
author: "SandraTietscher"
date: "2020-11-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```

This script looks in detail at selected cell type interactions

# Load libraries and data

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(SingleCellExperiment)
library(data.table)
library(dplyr)
library(magrittr)
library(dtplyr)
library(ggplot2)
library(gplots)
library(RColorBrewer)
library(tidyr)
library(ggpubr)
library(dittoSeq)

# Read SingleCellExperiment object
sce_protein <- readRDS("output/ProteinPanel/sce_ProteinPanel_subtypes_all.rds")

# Read object relationship file (output from Cellprofiler)
dat_relation <- fread("data/cpout/Object_relationships_ProteinPanel.csv")
dat_relation$cellA <- paste0(dat_relation$`First Image Number`, "_", dat_relation$`First Object Number`)
dat_relation$cellB <- paste0(dat_relation$`Second Image Number`, "_", dat_relation$`Second Object Number`)

# ROI info
ROI_info <- read.csv("data/ROI_info_HIER_TLSrevised.csv")
ROI_info$ImageNumber <- c(1:77)

#TLS image numbers
nonTLS_nr <- filter(ROI_info, TLS == "no")$ImageNumber
TLS_nr <- filter(ROI_info, TLS %in% c("pre", "yes"))$ImageNumber

TIG2 <- c("TBB111", "TBB129", "TBB165", "TBB171", "TBB184", "TBB338")
```

# Likelihood of PD1-high T cells neighbouring tumor cells

For TIG2 images only (TIG3 images often have so few PD-1 T cells that it distorts the image).

```{r PD1-tumor-neighbourhood, fig.height=4, fig.width=4}
cur_sce <- data.frame(colData(sce_protein))
cur_sce <- cur_sce[cur_sce[,"sample"] %in% TIG2,]
df <- data.frame(matrix(nrow=0, ncol =3))

for (i in unique(cur_sce$ImageNumber)){
  cur_sce_i <- cur_sce[cur_sce[,"ImageNumber"] == i,]
  
  # get T cell IDs
  cellIDs_PD1 <- cur_sce[cur_sce_i[,"subtype"] %in% c("CD4_PD1", "CD8_PD1"),]
  PD1_shortID <- paste0(cellIDs_PD1$ImageNumber, "_", cellIDs_PD1$CellNumber)
  cellIDs_nonPD1 <- cur_sce[cur_sce_i[,"subtype"] %in% c("CD4", "CD8", "Treg"),]
  nonPD1_shortID <- paste0(cellIDs_nonPD1$ImageNumber, "_", cellIDs_nonPD1$CellNumber)
  
  # get tumor cell IDs
  cellIDs_tumor <- cur_sce[cur_sce_i[,"celltype"] == "tumor",]
  tumor_shortID <- paste0(cellIDs_tumor$ImageNumber, "_", cellIDs_tumor$CellNumber)
  
  # subset neighbourhood list to include only `PD1-high T cell <-> tumor cell` interactions
  dat_relation_sub_PD1 <- dat_relation %>% filter((cellA %in% PD1_shortID & cellB %in% tumor_shortID) | (cellB %in% PD1_shortID & cellA %in% tumor_shortID))
  # extract PD1 IDs that border tumor cells
  PD1_tumor_bordering <- setdiff(unique(c(dat_relation_sub_PD1$cellA, dat_relation_sub_PD1$cellB)), tumor_shortID)
  PD1_perc <- length(PD1_tumor_bordering)/length(PD1_shortID)
  
  # subset neighbourhood list to include only `PD1-low T cell <-> tumor cell` interactions
  dat_relation_sub_nonPD1 <- dat_relation %>% filter((cellA %in% nonPD1_shortID & cellB %in% tumor_shortID) | (cellB %in% nonPD1_shortID & cellA %in% tumor_shortID))
  # extract PD1 IDs that border tumor cells
  nonPD1_tumor_bordering <- setdiff(unique(c(dat_relation_sub_nonPD1$cellA, dat_relation_sub_nonPD1$cellB)), tumor_shortID)
  nonPD1_perc <- length(nonPD1_tumor_bordering)/length(nonPD1_shortID)
  
  new_row <- c(i, PD1_perc, nonPD1_perc)
  df <- rbind(df, new_row)
} 

colnames(df) <- c("sample", "PD1_perc", "nonPD1_perc")
df[,2:3] <- df[,2:3] %>% mutate_if(is.character,as.numeric)

# Plot
ggpaired(df, cond1 = "PD1_perc", cond2 = "nonPD1_perc", line.color = "gray", ylab = "Direct tumor neighbour (proportion)", xlab = FALSE)+
  stat_compare_means(paired = TRUE)
```

### CD8 only, TIG2 only

```{r PD1-CD4-tumor-neighbourhood}
cur_sce <- data.frame(colData(sce_protein))
cur_sce <- cur_sce[cur_sce[,"sample"] %in% TIG2,]
df <- data.frame(matrix(nrow=0, ncol =3))

for (i in unique(cur_sce$ImageNumber)){
  cur_sce_i <- cur_sce[cur_sce[,"ImageNumber"] == i,]
  
  # get T cell IDs
  cellIDs_PD1 <- cur_sce[cur_sce_i[,"subtype"] == "CD4_PD1",]
  PD1_shortID <- paste0(cellIDs_PD1$ImageNumber, "_", cellIDs_PD1$CellNumber)
  cellIDs_nonPD1 <- cur_sce[cur_sce_i[,"subtype"] %in% c("CD4"),]
  nonPD1_shortID <- paste0(cellIDs_nonPD1$ImageNumber, "_", cellIDs_nonPD1$CellNumber)
  
  # get tumor cell IDs
  cellIDs_tumor <- cur_sce[cur_sce_i[,"celltype"] == "tumor",]
  tumor_shortID <- paste0(cellIDs_tumor$ImageNumber, "_", cellIDs_tumor$CellNumber)
  
  # subset neighbourhood list to include only `PD1-high T cell <-> tumor cell` interactions
  dat_relation_sub_PD1 <- dat_relation %>% filter((cellA %in% PD1_shortID & cellB %in% tumor_shortID) | (cellB %in% PD1_shortID & cellA %in% tumor_shortID))
  # extract PD1 IDs that border tumor cells
  PD1_tumor_bordering <- setdiff(unique(c(dat_relation_sub_PD1$cellA, dat_relation_sub_PD1$cellB)), tumor_shortID)
  PD1_perc <- length(PD1_tumor_bordering)/length(PD1_shortID)
  
  # subset neighbourhood list to include only `PD1-low T cell <-> tumor cell` interactions
  dat_relation_sub_nonPD1 <- dat_relation %>% filter((cellA %in% nonPD1_shortID & cellB %in% tumor_shortID) | (cellB %in% nonPD1_shortID & cellA %in% tumor_shortID))
  # extract PD1 IDs that border tumor cells
  nonPD1_tumor_bordering <- setdiff(unique(c(dat_relation_sub_nonPD1$cellA, dat_relation_sub_nonPD1$cellB)), tumor_shortID)
  nonPD1_perc <- length(nonPD1_tumor_bordering)/length(nonPD1_shortID)
  
  new_row <- c(i, PD1_perc, nonPD1_perc)
  df <- rbind(df, new_row)
} 

colnames(df) <- c("sample", "PD1_perc", "nonPD1_perc")
df[,2:3] <- df[,2:3] %>% mutate_if(is.character,as.numeric)

# Plot
ggpaired(df, cond1 = "PD1_perc", cond2 = "nonPD1_perc", line.color = "gray", ylab = "Direct tumor neighbour (proportion)", xlab = FALSE)+
  stat_compare_means(paired = TRUE)
```


# Likelihood of PD1-high T cells neighbouring aDCs

For all images.

```{r PD1-aDC-neighbourhood, fig.height=4, fig.width=4}
cur_sce <- data.frame(colData(sce_protein))
df <- data.frame(matrix(nrow=0, ncol =3))

for (i in unique(cur_sce$sample)){
  cur_sce_i <- cur_sce[cur_sce[,"sample"] == i,]
  
  # get T cell IDs
  cellIDs_PD1 <- cur_sce[cur_sce_i[,"subtype"] %in% c("CD4_PD1", "CD8_PD1"),]
  PD1_shortID <- paste0(cellIDs_PD1$ImageNumber, "_", cellIDs_PD1$CellNumber)
  cellIDs_nonPD1 <- cur_sce[cur_sce_i[,"subtype"] %in% c("CD4", "CD8", "Treg"),]
  nonPD1_shortID <- paste0(cellIDs_nonPD1$ImageNumber, "_", cellIDs_nonPD1$CellNumber)
  
  # get aDC IDs
  cellIDs_aDC <- cur_sce[cur_sce_i[,"celltype"] == "aDC",]
  aDC_shortID <- paste0(cellIDs_aDC$ImageNumber, "_", cellIDs_aDC$CellNumber)
  
  # subset neighbourhood list to include only `PD1-high T cell <-> aDC` interactions
  dat_relation_sub_PD1 <- dat_relation %>% filter((cellA %in% PD1_shortID & cellB %in% aDC_shortID) | (cellB %in% PD1_shortID & cellA %in% aDC_shortID))
  # extract PD1 IDs that border aDCs
  PD1_aDC_bordering <- setdiff(unique(c(dat_relation_sub_PD1$cellA, dat_relation_sub_PD1$cellB)), aDC_shortID)
  PD1_perc <- length(PD1_aDC_bordering)/length(PD1_shortID)
  
  # subset neighbourhood list to include only `PD1-low T cell <-> aDC` interactions
  dat_relation_sub_nonPD1 <- dat_relation %>% filter((cellA %in% nonPD1_shortID & cellB %in% aDC_shortID) | (cellB %in% nonPD1_shortID & cellA %in% aDC_shortID))
  # extract PD1 IDs that border aDCs
  nonPD1_aDC_bordering <- setdiff(unique(c(dat_relation_sub_nonPD1$cellA, dat_relation_sub_nonPD1$cellB)), aDC_shortID)
  nonPD1_perc <- length(nonPD1_aDC_bordering)/length(nonPD1_shortID)
  
  new_row <- c(i, PD1_perc, nonPD1_perc)
  df <- rbind(df, new_row)
} 

colnames(df) <- c("sample", "PD1_perc", "nonPD1_perc")
df[,2:3] <- df[,2:3] %>% mutate_if(is.character,as.numeric)
df$TIG <- ifelse(df$sample %in% TIG2, "TIG2", "TIG3")

# Plot
ggpaired(df, cond1 = "PD1_perc", cond2 = "nonPD1_perc", line.color = "gray", ylab = "Direct aDC neighbour (proportion)", xlab = FALSE)+
  stat_compare_means(paired = TRUE)


```

# T cell suppressive signature


```{r myeloid-suppressive-TcellSignature}
sce_myeloid <- sce_protein[,which(sce_protein$celltype == "myeloid")]

# Define "suppressive" myeloid cells
suppr_counts <- data.frame(t(assay(sce_myeloid, "counts")[c("CD274", "IDO1"),]))
sce_myeloid$suppressive <- ifelse(suppr_counts$CD274 > 1 | suppr_counts$IDO1 > 4, TRUE, FALSE)
prop.table(table(sce_myeloid$suppressive))

# Plot IDO1 vs CD274 expression & choose cutoffs
dittoScatterPlot(sce_myeloid, "IDO1", "CD274", color.var = "suppressive")+
  geom_hline(yintercept = 1, lty = "dashed")+
  geom_vline(xintercept = 4, lty = "dashed")

# ROI-wide correlation with PD1 T cells (excluding Tregs)
ROI_myeloid <- prop.table(table(sce_myeloid$ImageNumber, sce_myeloid$suppressive), margin = 1)[,2]
sce_T <- sce_protein[,which(sce_protein$subtype %in% c("CD4", "CD8", "CD4_PD1", "CD8_PD1"))]
sce_T$PD1 <- ifelse(sce_T$subtype %in% c("CD4", "CD8"), FALSE, TRUE)
ROI_T <- prop.table(table(sce_T$ImageNumber, sce_T$PD1), margin = 1)[,2]
ROI_comb <- as.data.frame(cbind(ROI_myeloid, ROI_T))
ROI_comb$TIG <- ROI_info$TIG
ROI_comb$sample <- ROI_info$sample

ggplot(ROI_comb, aes(ROI_myeloid, ROI_T))+
  geom_point(aes(color = sample))+
   theme(panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank())+
  ylab("Proportion of PD-1 high T cells\n(out of all T cells, excl. Tregs)")+
  xlab("Proportion of suppressive myeloid cells (high in PD-L1 and/or IDO1)\n(out of all myeloid cells)")+
  stat_cor(method="spearman", label.x.npc = "middle")+
  geom_smooth(method = lm, color = "black", se=FALSE)
```

## PD-1 status of the 10 closest T cells (for suppressive and non-suppressive myeloid cells)

Exclude Tregs.

```{r suppr-myeloid-PD1high-neighbours, fig.width=3.5, fig.height=4}
cur_sce <- data.frame(colData(sce_protein))
cellIDs_T <- cur_sce[cur_sce[,"subtype"] %in% c("CD4", "CD8", "CD4_PD1", "CD8_PD1"),]
cellIDs_PD1low <- rownames(cur_sce[cur_sce[,"subtype"] %in% c("CD4", "CD8"),])
cellIDs_T$PD1 <- ifelse(rownames(cellIDs_T) %in% cellIDs_PD1low, FALSE, TRUE)
new_sce <- data.frame(matrix(ncol=length(colnames(cur_sce))+1, nrow = 0))

# For each image separately: nearest neighbour search

for (j in unique(cur_sce$ImageNumber)) {
  cur_sce_j <- filter(cur_sce, ImageNumber == j)
  T_j <- filter(cellIDs_T, ImageNumber == j)[,c("Center_X", "Center_Y")]
  T_j_PD1 <- filter(cellIDs_T, ImageNumber == j)[,c("PD1")]
  
  # Search 10 closest T cells for each cell
    j_coord <- cur_sce_j[,c("Center_X", "Center_Y")]
    nn <- RANN::nn2(T_j, j_coord)
    
  # Check PD-1 status for each neighbouring T cell and calculate PD1-high proportions
    nn$nn.idx.PD1 <- t(apply(nn$nn.idx, 1, function(x) T_j_PD1[x]))
    nn$nn.PD1low.prop <- apply(nn$nn.idx.PD1, 1, function(x) prop.table(table(x))[1])
    cur_sce_j$neigh_PD1 <- 1 - nn$nn.PD1low.prop 

  # Add to full sce
    new_sce <- rbind(cur_sce_j, new_sce)
}

# Add T_distances to colData of original object
sce_protein$neigh_PD1 <- new_sce[order(match(new_sce$cellID, sce_protein$cellID)),]$neigh_PD1

# Subset myeloid cells and define suppressive status as above
sce_myeloid <- sce_protein[,which(sce_protein$celltype == "myeloid")]
suppr_counts <- data.frame(t(assay(sce_myeloid, "counts")[c("CD274", "IDO1"),]))
sce_myeloid$suppressive <- ifelse(suppr_counts$CD274 > 1 | suppr_counts$IDO1 > 4, TRUE, FALSE)

# Plot neighbouring PD1-Tcell-proportion for suppressive vs non-suppressive myeloid cells
myeloid_coldata <- data.frame(colData(sce_myeloid))

ggplot(myeloid_coldata, aes(x= suppressive, y = neigh_PD1))+
  geom_boxplot(outlier.shape = NA)+
  theme(axis.text.x=element_text(angle = 90, hjust=1),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank())+
  ylab("PD1-high T cell proportion\n(considering 10 closest T cells)")+
  xlab("Myeloid cells: T-suppressive signature")+
  stat_compare_means(method = "wilcox.test", aes(label=..p.signif..), label.x.npc = 0.5, label.y.npc = 0.8)
```



