---
title: "XX_RNApanel_cluster_neighbourhood"
author: "SandraTietscher"
date: "2021-02-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

This script calculates whether cytokine-expressing cells form clusters more than would be expected randomly.

# Preparations

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message= FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Load libraries

First, we will load the libraries needed for this part of the analysis.

```{r load-libraries, message=FALSE}
sapply(list.files("code/helper_functions", full.names = TRUE), source)
library(SingleCellExperiment)
library(reshape2)
library(tidyverse)
library(dplyr)
library(data.table) 
library(ComplexHeatmap)
library(circlize)
library(gridExtra)
library(neighbouRhood)
library(viridis)
```

## Load CP output and SCE

```{r}
# Load and prepare
sce_rna = readRDS(file = "output/RNApanel/sce_RNApanel_cytokines_communities.rds")
dat_cells = fread(file = "~/home//Segmentation/BCexh_RNAscope/cpout/cell.csv",stringsAsFactors = FALSE)
dat_relation = fread("data/cpout/Object_relationships_RNApanel.csv",stringsAsFactors = FALSE)

# define number of permutations
n_perm = 100 
```

## Start the analysis (takes ~30 min if run on one core only)

```{r milieu_analysis, eval=F}
#start = Sys.time()
sce_adapt <- sce_rna
sce_adapt$cellID <- paste0("R_", sce_adapt$ImageNumber, "_", sce_adapt$CellNumber)
cur_sce <- as.data.frame(colData(sce_adapt))

TIG_images <- as.data.frame(table(sce_adapt$ImageNumber, sce_adapt$TIG))
TIG_images <- TIG_images[which(TIG_images$Freq > 0),]
colnames(TIG_images) <- c("ImageNumber", "TIG", "cells")

# add same cellID to dat_cells as in sce object
dat_cells$cellID <- paste("R_", paste(dat_cells$ImageNumber, dat_cells$ObjectNumber, sep = "_"), sep = "")

size <- 3
image_df <- data.frame()
for(j in c("TIG2", "TIG3")) {
  selected_images <- as.vector(filter(TIG_images, TIG == j)[,1])
  images <- data.frame()
  # loop through the colnames indicating if a cell is positive for a certain chemokine
  for(i in colnames(cur_sce[,grepl("CCL|CXCL|CSF",colnames(cur_sce))])){
    # add chemokine info to celltype
    sce_info <- cur_sce[,c("cellID", i , "celltype")]
    
    # add celltype information
    dat_cells_tmp <- left_join(as.data.frame(dat_cells), sce_info, by = "cellID")
    
    #assign labels and groups
    dat_cells_tmp$label <- dat_cells_tmp[,i]
    dat_cells_tmp$group <- dat_cells_tmp$celltype
    dat_cells_tmp <- as.data.table(dat_cells_tmp)
    
    # subset dat_relation and dat_cells
    dat_cells_sub <- dat_cells_tmp#[dat_cells$celltype == "P3",]
    dat_relation_sub <- dat_relation[which(dat_relation$`First Image Number` %in% selected_images),]
    
    # Prepare the data
    d = neighbouRhood::prepare_tables(dat_cells_sub, dat_relation_sub)
    
    # Calculate the baseline statistics
    dat_baseline = neighbouRhood::apply_labels(d[[1]], d[[2]]) %>%
      neighbouRhood::aggregate_classic_patch(., patch_size = size)
    
    # Calculate the permutation statistics
    # This will run the test using parallel computing. The name of the idcol does actually not matter.
    
    set.seed(12312)
    dat_perm = rbindlist(lapply(1:n_perm, function(x){
      dat_labels = neighbouRhood::shuffle_labels(d[[1]])
      neighbouRhood::apply_labels(dat_labels, d[[2]]) %>%
        neighbouRhood::aggregate_classic_patch(., patch_size = size)
    }
    ), idcol = 'run')
    
    # calc p values
    dat_p <- neighbouRhood::calc_p_vals(dat_baseline, dat_perm, n_perm = n_perm, p_tresh = 0.01) 
    
    # select interactions between chemokine+ cells
    dat_p$interaction <- paste(dat_p$FirstLabel, dat_p$SecondLabel, sep = "_")
    
    dat_p_wide <- dat_p %>%
      reshape2::dcast(group ~ interaction, value.var = "sigval", fill = 0) %>%
      select(group, `1_1`)
    
    summary <- as.data.frame(dat_p_wide) %>%
      group_by(`1_1`) %>%
      summarise(n=n(),.groups = 'drop') %>%
      ungroup() %>%
      mutate(percentage_sig = (n/sum(n)) * 100)
    
    images <- rbind(images, cbind(summary[1,], i))
  }
  
  # calculate percentage of images with significant patches
  images$percentage_sig <- 100 - images$percentage_sig
  images$patch_size <- size
  images$TIG <- j
  images <- select(images, percentage_sig, i, patch_size, TIG)
  colnames(images) <- c("significant_images", "chemokine", "patch_size", "TIG")
  
  # add to data.frame
  image_df <- rbind(image_df, images)
}
#end = Sys.time()

#print(end-start)
```

# Plot the results

Plot percentage of images where 3-cell-patches of the given cytokine-expressing cells occur more often than randomly expected.

```{r patch-comparison-TIG, fig.width= 5, fig.height=3}
# image_df$chemokine <- gsub("_expression", "", image_df$chemokine)
# 
# ggplot(image_df, aes(x = chemokine, y = significant_images, fill = TIG))+
#   geom_bar(position="dodge", stat="identity")+
#   theme(panel.border = element_rect(color = "black", fill = NA, size = 1),
#         panel.background = element_blank(), 
#         axis.text.x = element_text(angle = 60, hjust = 1))
```

